### 2017-10-10 response/response-jde-1.tex

[RESPONSE:] We have changed the equation and the associated text.

\item What other time-varying covariates are available in the survey? It
would be interesting to consider the extent to which the crop loss
results are explained by other household-level idiosyncratic variation.

[RESPONSE:]

There are three time-varying covariates in the survey that could 
potentially affect crop loss or the outcome variables directly:
rainfall, prices, and loss of livestock.
Other time-varying variables are illness and health variables, which 
we already show are unlikely to explain the results we find.

@@

As we discuss in detail below in response to Reviewer 2, higher 
rainfall is associated with lower risk of crop loss, although there
are a number of caveats to that result.
Furthermore, including rainfall as an independent covariate does 
not change the relationship between crop loss and the five main outcomes.

In Appendix Table A-14 we show the main results when including a dummy for livestock
loss and the price index included with the Kagera data.
Although both new variables generally have the expected sign, neither are 
close to statistically significant and there is only minimal change in the 
estimated coefficients for crop loss.
There are two caveats to the results.
First, the loss of livestock question is not as ``clean'' as the
crop loss question in that the livestock question also includes 
any livestock given away by the household without specifying how much 
is actual loss and how much is gifts.
Second, the prices used for the price index are measured at the time
of the survey.
We tried lagging the price index, but that does not change the findings.

### 2017-10-10 response/response-jde-1.tex

% - rainfall does not do anything
% - illness of mother and other household members

\item The authors report some nonsensical coefficients and standard
errors like -0.000 (0.000). All coefficients and standard errors should
be written to two significant digits.

[RESPONSE:] The 0.000 coefficients were predominantly for variables
involving initial assets. 
We have changed the variable definition for assets, so that it is
now measured in 1,000,000 TZS.

Note that there are still a few places---mainly for estimations
involving the use of modern contraceptives---where the estimated 
coefficients are so small that presenting additional digits does not make
much economic sense.
In those cases the standard errors are always equal to or larger than
the coefficients.
See, for example, Table A-10 on the effects of log crop loss.

@@

\item Table 6 regresses pregnancy on contraceptive use, crop loss, and
their interaction. As the authors note, this regression is quite
difficult to interpret, given the endogeneity of contraceptive use. I
think this table adds little to the paper.

[RESPONSE:] Although we understand the concern about endogeneity, 
we believe the table is worth keeping. 
In the demography literature there is a substantial disagreement on whether
sustained fertility reduction requires access to modern contraceptives.
Part of the problem in this literature is that the ``intensity'' with
which a couple uses contraceptives---either traditional or modern---is
difficult to measure correctly.
% [something on the panel nature of the data and that that is important
% for getting at the incentives and the use over time; not perfect but
% at least is for a shorter period than other (cross-sectional) surveys]
The table is important because it provides us a way to get at this
intensity by showing that only couples that have a strong incentive to
prevent pregnancies appear to be successful users of contraceptives.
Furthermore, this shows that it is possible to postpone pregnancies with
just traditional contraceptives because traditional contraceptives account 
for the vast proportion of contraception used.

% [also speaks to whether this is a conscious decision or not] 
The table also supports our conjecture that the estimated effects 
% [might need a less causal word] 
of crop loss capture a conscious decision. 
If shocks lead to lower probability of pregnancy and birth because
of other factors than the decision to use contraceptive we should see
a negative and statistically significant coefficient for the crop loss
dummy by itself.
We do not.
Instead, what we get is a substantial negative coefficient for the interaction
between contraceptive use and crop loss and a coefficient for crop loss 
on its own that is positive and far from statistically significant.

### 2017-10-10 response/response-jde-1.tex

\begin{document}

\begin{center} \textbf{\large Income Shocks, Contraceptive Use, and
Timing of Fertility} \end{center}

\begin{center} Response to Referee Comments \end{center}

\noindent Please find attached the revised version of our paper,
``Income Shocks, Contraceptive Use, and Timing of Fertility.''
We would like to thank the two referees for the very useful comments and
suggestions,
which we believe have improved the paper.

@@

We rewrote the code for cleaning and analyzing the data from the ground up 
to make it easier to follow and eventually publish on GitHub.
This rewrite lowered the sample by three observations.
Including all married/partnered women up to age 45, rather than exclude
women younger than 18, the final sample size is, however, the same as 
before and the results are essentially unchanged.

To remain within the page limit we had to substantially reorganize
part of the paper.
The following is a list of the more substantial changes made:
\begin{enumerate}
\item We removed the map of Kagera.
\item In response to the reviewers' comments we have almost completely
rewritten the Robustness section as detailed below.
\item As part of the rewrite of the Robustness section we moved 
the discussion of age, wealth, and education results to a new Appendix
section since these results do little to clarify the mechanisms 
behind the main results.
We still address and incorporate the comments from reviewers that
relate to these outcomes.
\item We moved the discussion of estimation strategy for age,
wealth, and education estimation to the new Appendix section.
\item We have added another new Appendix section that examine results 
with including an additional lag for crop loss (see response to
Reviewer 1 below).
\item A number of new Appendix tables have been added in response
to various comments.
\end{enumerate}

@@

Although the results are potentially interesting in themselves,
because of data limitation the approach unfortunately does not  
provide a clean test for whether households postpone or reduce their 
fertility in response to shocks.
First, given the short duration between surveys and the lack of 
information about when the shocks occurred, it is possible and
even likely that the effects of a crop loss last for more than one period.
Hence, if we find a negative effect of more distant crop losses
this may be the result of the negative effects of the crop losses
still affecting the household.
Second, we lose 1/4 of the original number of observations each 
time we include an additional lag, which leaves us with less variation 
in crop loss because the first round had a higher proportion of crop 
losses than the subsequent rounds.

Even if crop losses' impact are limited to only one period, there 
are two additional concerns.
First, the compensatory effect may be spread over multiple 
periods because of the uncertainty associated with waiting time
to conception, i.e. a couple may be trying to conceive in the
period after the crop loss but fail.
This makes a positive compensatory effect harder to 
identify.
Second, it is possible that households may respond differently to a 
shock if they are hit repeatedly.
There are two competing factors at work.
On one hand, repeated shocks may exacerbate the negative impact on the
household, making it \emph{more} likely that fertility is postponed.
On the other hand, even longer spacing between births can be costly 
because economies of scale are harder to realize and because the 
mother may have to  be out of the labor force for longer if she 
already have young children to care for, which will make it
\emph{less} likely that fertility is postponed.
To capture the idea that repeated shocks have an additional
impact we therefore also include the interaction between 
crop losses.

With the data limitations and additional concerns,
it is difficult to draw strong predictions on what the coefficients
should be (except that the initial impact should lead to postponement 
of fertility).
Including only the additional lag of crop loss it appears that 
there is a weak positive association between crop loss in the prior
period and the likelihood of pregnancy and birth, although the 
coefficients are small and statistically insignificant.
There is still a negative effect of the immediate crop loss
variable, although these are also not statistically significant.
These results are broadly consistent with a compensatory response
where crop loss longer ago leads to higher fertility now.
However, if we include the interaction between crop losses we 
instead find negative association for both immediate and lagged 
crop loss with a positive interaction coefficient.
Even with the positive interaction coefficient the estimated 
impact of crop losses in the two prior periods is negative.
For contraceptive use we see similar patterns (with the signs 
reversed).

@@

% [also speaks to whether this is a conscious decision or not] 
The table also supports our conjecture that the estimated effects 
% [might need a less causal word] 
of crop loss capture a conscious decision. 
If shocks lead to lower probability of pregnancy and birth because
of other factors than the decision to use contraceptive we should see
a negative and statistically significant coefficient for the crop loss
dummy by itself.
We do not.
Instead, what we get is a substantial negative coefficient for the interaction
between contraceptive use and crop loss and a coefficient for crop loss 
on its own that is positive and far from statistically significant.

\item Several appendix tables interact crop loss with initial assets in
Round 1 of the survey. This exercise is interesting and worthy of doing,
given that one might expect asset-poor households to have particular
difficulty in smoothing consumption. But the authors should leave Round
1 outcomes out of this regression.

[RESPONSE:] We have added the results where round 1 shocks has been 
excluded to the relevant table in the Appendix section on age
groups, wealth, and education.
The results are broadly consistent with what we originally found
for pregnancy and birth, 
The contraceptive use results have the correct sign, but the direct
effect is not statistically significant, presumably because of the 
lower variation in crop loss and the smaller sample size.

### 2017-10-10 paper/incomeShocks-jde-r1.tex
Minor edits
Minor edits to shorten paper

% Income Shocks, Contraceptive Use, and Timing of Fertility
% incomeShocks-jde-r1.tex
% Begun.: 2017-07-13
% Edited: 2017-07-13

% Revisions based on Demography referee reports

@@

\date{September 2017}

@@

How shocks affect fertility decisions in developing countries has 
received little attention in the literature.
A correlation between shocks and the timing of births is well-established, but it is 
unclear whether the cause is households' intentional planning or unintended consequences 
of the shocks \citep{pitt98b,lindstrom99,Evans2010,Portner2014}.%
\footnote{
There is a related literature on the effects of weather, especially temperature,
on timing of fertility \citep[see, for example,][]{Barreca2015}.
%  and TK Josh paper.
In a different vein, a monthlong blackout in Zanzibar increased 
births 8 to 10 months later \citep{Burlando2014}.
There is also a large literature examining how households in developing countries
cope with shocks in areas such as labor supply, crop choice, and education.
}
Furthermore, there is, in general, little evidence on what determines timing 
of fertility in developing countries, with most of the literature instead 
focusing on the number of children born \citep{Portner2018}.
% [Little evidence in general on timing of fertility in developing countries;
% most of the research has focused on completed fertility / children-ever-born] 
% Furthermore, standard models of fertility focus mainly on the effects of time
% invariant household and individual characteristics on fertility outcomes, and 
% tend to ignore the dynamic aspects of fertility decisions.%
% \footnote{
% For an early exception see \cite{Newman1984}.
% }

@@

Observing a decline in pregnancies and birth rates following a shock is, however, 
not evidence of a conscious decision to change the timing of fertility.
A shock may, for example, lead one or more members of the household to 
migrate in search of better economic opportunities, and if either the husband 
or the wife are gone for extended periods, this will lower the probability of 
conception.
Furthermore, a shock may reduce intercourse frequency---either for
physiological or psychological reasons or from increased hours of work.
We would also observe a decline in births if people delay marriage or 
if there is an increase in the dissolution of marriage.
Finally, severe shocks can lead to starvation and health problems \citep{lindstrom99}.
Malnourished pregnant mothers are more likely to have stillbirths,
and non-pregnant women may see an increased incidence of secondary sterility
from a reduction in age of menopause or famine induced amenorrhea.%
\footnote{
See \cite{Hernandez-Julian2014} for a review of the literature and an 
analysis of the effect of famine on stillbirths.
}

@@

Most of the prior literature on the impact of economic and other shocks on fertility 
has used historic data on what are now developed countries.
Data from Rouen, France, over the period 1681-1787, show that increases in the
price of wheat led to a dramatic fertility decline for the urban poor, 
whereas fertility of the urban wealthy was unaffected \citep{Galloway1987}.
Similarly, English data from 1542 to 1800 show that mortality shocks led to short-term
fertility declines, with the largest decline typically occurring  the year following 
the shock \citep{Bailey1998}.%
\footnote{
This data also indicated that an increase in real wage led to an increase in short-term 
fertility.
Increases in wage rate and declines in child mortality explain a significant part of the 
long-term fertility decline in Sweden during the period 1736 to 1946 \citep{Eckstein1999a}.
In another study on Sweden, increases in the value of women's time relative to men's 
time---arising from increases in the price of butter---led to a decline in fertility during 
the  1860-1910 period \citep{Schultz1985}. 
}
In pre-industrial Sweden, positive crop shock, positive weather shock, and 
positive wage shock all increased fertility for a five- to ten-year period following 
the shocks \citep{Eckstein1985}.
However, the increase in birth rate changed only the timing of births with no 
cumulative effect on the long-term fertility rate.
Similarly, the increasing price of rye in Southern Sweden led landless and 
semi-landless families to reduce their fertility, whereas landed farmers 
showed no change \citep{bengtsson06}.
Using present-day data, there is evidence that famine and domestic/regional 
military attacks in Ethiopia led to short-term decreases in the likelihood of 
conception \citep{lindstrom99}.
Closer to our paper, positive rainfall shocks in Senegal increased 
fertility \citep{pitt98b}.
None of these paper can, however, directly address whether the observed changes 
arise from deliberate control of fertility.

Many studies show how a variety of factors affect contraception use, such 
as, 
schooling, 
focused information campaigns, and
participation in savings or credit group 
\citep{Ainsworth1996,Feyisetan1996,Steele2001,Chen2003}.
To our knowledge, only one study attempts to show the effect of shocks on 
contraception use \citep{Hernandez-Correa2010}. 
Cross-sectional data from Madagascar show that households that experience adverse 
events are more likely to use contraception.\footnote{%
Adverse events include economic and environmental aspect, such as: rise in input 
cost, rise in cost of goods, difficulty finding buyers, difficulty finding inputs, 
floods, late rains, early rains, droughts, pest problems, etc. 
} 
However, 
% households may be able to anticipate many of these events before 
% they occur and adjust their behavior accordingly.
% Furthermore, 
the possible correlation between the probability of experiencing 
an adverse event and unobserved individual and household characteristics may 
bias the estimates, and the cross-sectional data allows for neither household 
nor individual level fixed effects estimates.

@@

On average across the four surveys, 17\% of women report being pregnant 
at the time of the interviews and 16\% of women report giving birth since the 
prior survey round.
In total, just below 70\% of women in the sample report giving birth in
the covered period (approximately 3\sfrac{1}{2} years).
Similarly, 70\% report a pregnancy at one point during the 4 rounds.

@@

% [Descriptive statistics and discussion here]
\input{../tables/desstat1.tex}

@@

A notable characteristic is the high level of reported education for a poor 
rural area, although there also appears to be errors in 
reported education levels.%
\footnote{
For many women, the reported education level decreases in certain survey 
rounds, or increases by 2 years or more in consecutive survey rounds, which 
conflicts with the 6 to 7 months survey intervals. 
Additionally, individuals are supposed to complete primary school by the age of 15,
so changes in primary schooling levels in our sample is likely from reporting 
errors. 
For women who report different primary school education levels, we 
assign the modal education level if possible, or, if there is no modal 
observation, the median years of education.
We use the same approach for household heads' education
when we use household heads' education as a proxy for wealth.
}
We divide education levels into three groups: no education, 
some primary (1-6 years of education), and primary school 
or above (7 or more years of education).
Only 29\% of the women received no education, 27\% received some 
primary education, and 44\% completed at least primary education.
The high reported education level is most likely the result of the 1974 
Universal Primary Education Movement,
which increased accessibility of primary education and enrollment rates, 
even though there are reports that the quality of education was very low.%
% In addition, the crisis Tanzania experienced in the 1980s further lowered the 
% quality and enrollments declined significantly.%
\footnote{
See \cite{Galabawa2001} and \cite{Wedgwood2005} for detailed discussions 
of the development of education in Tanzania.
}
Hence, it is unclear to what extent reported education levels reflect 
women's actual human capital.

@@

% [BASELINE estimations]

@@

Finally, we discuss a number of robustness checks for our results.
The most important concern is our definition of crop loss.
Above we define crop loss as a loss of more than 200 TZS.
To check how sensitive our results are to this cut-off,
we re-ran the regressions with varying cut-offs from 
0 to 5,000 TZS.%
\footnote{
The results are shown in Appendix Table \ref{tab:cut_offs}.
}
The signs of the coefficients are the same across all cut-offs.
Between 100 and 1,000 TZS, crop loss has a statistically significant
effect on same outcome as above.
Above 1,000 TZS the effects on pregnancy and birth become smaller and 
eventually statistically significant insignificant.
This is not surprising.
With a higher cut-off, more and more households will have experienced 
a crop loss but are not counted as having one, which will tend to 
bias results towards zero if crop loss have an effect at levels 
lower than the cut-off.%
\footnote{
This is supported by regressions---not shown---where we drop all observations 
with crop loss larger than zero but smaller or equal to the cut-off and find 
substantially larger coefficients.
} 
Using the log of crop loss instead of a cut-off does
not change our conclusion that crop loss significantly reduces
the likelihood of a pregnancy and birth and significantly
increases the use of contraception.%
\footnote{
See Appendix Table \ref{tab:ln_croploss}.
Using continuous crop loss leads to similar signed coefficients,
although neither pregnancy or birth are statistically significant
as shown in Appendix Table \ref{tab:continuous}.
Given the unattractive assumption that the effect of an unit
increase in crop loss is the same at all levels inherent in
this specification, the lack of statistically significance is
not surprising. 
}
The same is the case if we use a Logit model instead of the
linear probability model.%
\footnote{
See Appendix Table \ref{tab:logit}.
}

@@

% combine these two questions
% Omitted variables (based on R2 question)
% Time varying results (based on R1 question)
Furthermore, our definition of crop loss is based on the 
household's assessment and cannot be directly verified.
It is therefore possible that omitted, time-varying 
factors are responsible for the results. 
We approach this issue in two ways.
First, we included two other time-varying factors that
may affect fertility, loss of livestock and a price
index, in the main regressions.%
\footnote{
See results in Appendix Table \ref{tab:timevarying}.
Note, that the loss of livestock unfortunately also
includes livestock given away by the household.
}
Neither of these variables are statistically significant
and the coefficient on crop loss does not change.
[TK bla, bla - next sentences need a rewrite]
Although this does not completely rule out the possibility
that other time-varying factors, such as an epidemic or
local conflict, could affect fertility decisions, it does
imply that unobserved time-varying factors were not a 
major influence.
Conflict and epidemics would most likely have had 
substantial effects on either or both of the new
time-varying variables included.

@@

Although we cannot entirely rule out that other, unobserved
factors play a role, our main conclusion is that households appear
to consciously delay fertility through contraceptive use to cope 
with income shocks. 
This is especially of interest since this is accomplished 
predominately through the use of traditional contraceptives.
If traditional contraceptives have low efficacy, a natural question is 
how the substantial reductions in likelihood of births and pregnancy 
come about.
We present suggestive evidence that show that only those households
with a strong incentive, i.e. those exposed to crop loss, are 
effective users of traditional contraceptives.%
\footnote{
This is in line with the results in \cite{Rosenzweig1989},
where women with the highest opportunity cost of children
are also the most effective user of traditional contraceptives,
although the setting is very different.
}

There are three potential caveats to our results.
First, the sample size is small and the survey rounds
could have been further apart, but the longitudinal nature
of the data means that we can control for unobservable 
time-invariant individual and household characteristics.
Second, not surprisingly there appears to be measurement 
errors.
% Reported age and education levels are not always consistent
% across rounds, which is an indication that we probably also
% have measurement errors in the other variables, although it is
% harder to verify.
To the extent that these measurement errors are classical 
measurement errors, this biases our results towards zero,
making it harder to show an effect of crop loss.%
\footnote{
The problems with measurement error, combined with the small
sample size, may also explain why we do not find clear
differences in the response to crop loss across age groups,
education, and wealth levels.
}
Third, although we cover a number of possible explanations
for what determines crop loss, it is possible that there may 
be other unobserved, time-varying household characteristics 
that influence whether a household experience a crop loss.
If that is the case the causality from crop to fertility decisions
may not hold.

@@

% [Lead with this or the next sentence?]
% Our result that people are able to postpone fertility using mainly 
% traditional contraceptive do, however, have implications for our 
% understanding of long-term fertility.
% [Alternative lead]
What implications, if any, do our results then have for our understanding 
of long-term fertility?
% What we do not know is whether the psychic cost of using traditional 
% contraceptives is so high that although households are willing to use 
% them to deal with short-term shocks, they become difficult to use for 
% long-run moderation of fertility.
The economic literature on fertility transition in Europe and the US 
show a substantial decline in fertility and control of the timing of 
fertility before the introduction of modern contraceptives, although 
the methods used are generally not well documented.%
\footnote{
See, for example, \citet{Guinnane2011} on fertility reduction and 
\citet{Cinnirella2017} on spacing of birth.
On contraceptive methods used see, for example, \citet{Michael1976},
\citet{David1986}, and \citet{Santow1995}.
}
Furthermore, studies of the introduction of modern family planning, 
using individual level data, rarely find reductions in fertility larger 
than one child.%
\footnote{
See, for example, \citet{Portner2014a} 
and \citet{Portner2018} and the literature surveyed therein.
}
Both set of results indicate that households are able to 
limit long-term fertility using traditional contraceptives.
Our individual level data allow us to directly link the use of
traditional contraceptives and fertility outcomes.
We argue that, contrary to common perception, traditional
contraceptive methods have a high degree of efficacy when the
incentives to prevent pregnancy or birth are sufficiently high.
Because of the small data set we cannot, however, differentiate between
the different traditional methods, although---contrary to Europe 
and the US---withdrawal plays little or no role in this area.

@@

Including only the additional lag of crop loss it appears that 
there is a weak positive association between crop loss in the prior
period and the likelihood of pregnancy and birth, although the 
coefficients are small and statistically insignificant.
There is still a negative effect of the immediate crop loss
variable, although these are also not statistically significant.
These results are broadly consistent with a compensatory response
where crop loss longer occur leads to higher fertility now.
However, if we include the interaction between crop losses we 
instead find negative association for both immediate and lagged 
crop loss with a positive interaction coefficient.
Even with the positive interaction coefficient the estimated 
impact of crop losses in the two prior periods is negative.
For contraceptive use we see similar patterns (with the signs 
reversed).

>>>>

% Income Shocks, Contraceptive Use, and Timing of Fertility
% incomeShocks-jde-r1.tex

% Revisions based on Journal of Development Economics referee reports

@@

\date{October 2017}

@@

How shocks affect fertility decisions in developing countries has 
received little attention in the literature.
A correlation between shocks and the timing of births is well-established, but it is 
unclear whether the cause is households' intentional planning or unintended consequences 
of the shocks \citep{pitt98b,lindstrom99,Evans2010,Portner2014}.%
\footnote{
There is a related literature on the effects of weather, especially temperature,
on timing of fertility \citep[see, for example,][]{Barreca2015}.
In a different vein, a monthlong blackout in Zanzibar increased 
births 8 to 10 months later \citep{Burlando2014}.
There is also a large literature examining how households in developing countries
cope with shocks in areas such as labor supply, crop choice, and education.
}
Furthermore, there is, in general, little evidence on what determines timing 
of fertility in developing countries, with most of the literature instead 
focusing on the number of children born \citep{Portner2018}.
% Little evidence in general on timing of fertility in developing countries;
% most of the research has focused on completed fertility / children-ever-born
% Furthermore, standard models of fertility focus mainly on the effects of time
% invariant household and individual characteristics on fertility outcomes, and 
% tend to ignore the dynamic aspects of fertility decisions.%
% \footnote{
% For an early exception see \cite{Newman1984}.
% }

@@

A decline in pregnancies and birth rates following a shock is, however, 
not evidence of a conscious decision to change the timing of fertility.
A shock may, for example, lead one or more household members to 
migrate in search of better economic opportunities, and if either the husband 
or the wife are gone for extended periods, this will lower the probability of 
conception.
Furthermore, a shock may reduce intercourse frequency---either for
physiological or psychological reasons or from increased hours of work.
We would also observe a decline in births if people delay marriage or 
if there is an increase in the dissolution of marriage.
Finally, severe shocks can lead to starvation and health problems \citep{lindstrom99}.
Malnourished pregnant mothers are more likely to have stillbirths,
and non-pregnant women may see an increased incidence of secondary sterility
from a reduction in age of menopause or famine induced amenorrhea.%
\footnote{
See \cite{Hernandez-Julian2014} for a review of the literature and an 
analysis of the effect of famine on stillbirths.
}

@@

Most of the literature on the impact of economic and other shocks on fertility 
has used historic data on what are now developed countries.
Data from Rouen, France, over the period 1681-1787, show that increases in the
price of wheat led to a dramatic fertility decline for the urban poor, 
whereas fertility of the urban wealthy was unaffected \citep{Galloway1987}.
Similarly, English data from 1542 to 1800 show that mortality shocks led to short-term
fertility declines, with the largest decline typically occurring  the year following 
the shock \citep{Bailey1998}.%
\footnote{
This data also indicated that an increase in real wage led to an increase in short-term 
fertility.
Increases in wage rate and declines in child mortality explain a significant part of the 
long-term fertility decline in Sweden during the period 1736 to 1946 \citep{Eckstein1999a}.
In another study on Sweden, increases in the value of women's time relative to men's 
time---arising from increases in the price of butter---led to a decline in fertility during 
the  1860-1910 period \citep{Schultz1985}. 
}
In pre-industrial Sweden, positive crop shock, positive weather shock, and 
positive wage shock all increased fertility for a five- to ten-year period following 
the shocks \citep{Eckstein1985}.
However, the increase in birth rate changed only the timing of births with no 
cumulative effect on the long-term fertility rate.
Similarly, the increasing price of rye in Southern Sweden led landless and 
semi-landless families to reduce their fertility, whereas landed farmers 
showed no change \citep{bengtsson06}.
Using present-day data, famine and domestic/regional military attacks in Ethiopia 
led to short-term decreases in the likelihood of conception \citep{lindstrom99}.
Closer to our paper, positive rainfall shocks in Senegal increased 
fertility \citep{pitt98b}.
None of these paper can, however, directly address whether the observed changes 
arise from deliberate control of fertility.

Many studies show how a variety of factors affect contraception use, such 
as, 
schooling, 
focused information campaigns, and
participation in savings or credit group 
\citep{Ainsworth1996,Feyisetan1996,Steele2001,Chen2003}.
To our knowledge, only one study attempts to show the effect of shocks on 
contraception use \citep{Hernandez-Correa2010}. 
Cross-sectional data from Madagascar show that households that experience adverse 
events are more likely to use contraception.\footnote{%
Adverse events include economic and environmental aspect, such as: rise in input 
cost, rise in cost of goods, difficulty finding buyers, difficulty finding inputs, 
floods, late rains, early rains, droughts, pest problems, etc. 
} 
However, 
% households may be able to anticipate many of these events before 
% they occur and adjust their behavior accordingly.
% Furthermore, 
the possible correlation between the probability of experiencing 
an adverse event and unobserved characteristics may bias the estimates, 
and the cross-sectional data allow for neither household 
nor individual level fixed effects estimates.

@@

On average, 17\% of women report being pregnant at the time of the interviews 
and 16\% of women report giving birth since the prior survey round.
In total, just below 70\% of women in the sample report giving birth in
the covered period (approximately 3\sfrac{1}{2} years).
Similarly, 70\% report a pregnancy at one point during the 4 rounds.

@@

% Descriptive statistics and discussion here
\input{../tables/desstat1.tex}

@@

A notable characteristic is the high level of reported education for a poor 
rural area, although there also appears to be errors in 
reported education levels.%
\footnote{
For many women, the reported education level decreases in certain survey 
rounds, or increases by 2 years or more in consecutive survey rounds, which 
conflicts with the 6 to 7 months survey intervals. 
Additionally, individuals are supposed to complete primary school by the age of 15,
so changes in primary schooling levels in our sample is likely from reporting 
errors. 
For women who report different primary school education levels, we 
assign the modal education level if possible, or, if there is no modal 
observation, the median years of education.
We use the same approach for household heads' education
when we use household heads' education as a proxy for wealth.
}
We divide education levels into three groups: no education, 
some primary (1-6 years of education), and primary school 
or above (7 or more years of education).
Only 29\% of the women received no education, 27\% received some 
primary education, and 44\% completed at least primary education.
The high education level is most likely the result of the 1974 
Universal Primary Education Movement,
which increased accessibility of primary education and enrollment rates, 
even though there are reports that the quality of education was very low.%
% In addition, the crisis Tanzania experienced in the 1980s further lowered the 
% quality and enrollments declined significantly.%
\footnote{
See \cite{Galabawa2001} and \cite{Wedgwood2005} for detailed discussions 
of the development of education in Tanzania.
}
Hence, it is unclear whether reported education levels reflect 
women's actual human capital.

@@

% BASELINE estimations

@@

Finally, we discuss a number of robustness checks for our results.
The most important concern is our definition of crop loss.
Above we define crop loss as a loss of more than 200 TZS.
To check how sensitive our results are to this cut-off,
we re-ran the regressions with varying cut-offs from 
0 to 5,000 TZS.%
\footnote{
The results are shown in Appendix Table \ref{tab:cut_offs}.
}
The signs of the coefficients are the same across all cut-offs.
Between 100 and 1,000 TZS, crop loss has a statistically significant
effect on same outcome as above.
Above 1,000 TZS the effects on pregnancy and birth become smaller and 
eventually statistically insignificant.
This is not surprising.
With a higher cut-off, more and more households will have experienced 
a crop loss but are not counted as having one, which will tend to 
bias results towards zero if crop loss have an effect at levels 
lower than the cut-off.%
\footnote{
This is supported by regressions---not shown---where we drop all observations 
with crop loss larger than zero but smaller or equal to the cut-off and find 
substantially larger coefficients.
} 
Using the log of crop loss instead of a cut-off does
not change our conclusion that crop loss significantly reduces
the likelihood of a pregnancy and birth and significantly
increases the use of contraception.%
\footnote{
See Appendix Table \ref{tab:ln_croploss}.
Using continuous crop loss leads to similar signed coefficients,
although neither pregnancy or birth are statistically significant
as shown in Appendix Table \ref{tab:continuous}.
Given the unattractive assumption that the effect of an unit
increase in crop loss is the same at all levels inherent in
this specification, the lack of statistically significance is
not surprising. 
}
The same is the case if we use a Logit model instead of the
linear probability model.%
\footnote{
See Appendix Table \ref{tab:logit}.
}

@@

% combine these two questions
% Omitted variables (based on R2 question)
% Time varying results (based on R1 question)
Furthermore, our definition of crop loss is based on the 
household's assessment and cannot be directly verified.
It is therefore possible that omitted, time-varying 
factors are responsible for the results. 
We approach this issue in two ways.
First, we included two other time-varying factors that
may affect fertility, loss of livestock and a price
index, in the main regressions.%
\footnote{
See results in Appendix Table \ref{tab:timevarying}.
Note, that the loss of livestock unfortunately also
includes livestock given away by the household.
}
Neither of these variables are statistically significant
and the coefficient on crop loss does not change.
Although this does not completely rule out the possibility
that other time-varying factors, such as an epidemic or
local conflict, could affect fertility decisions, it does
imply that unobserved time-varying factors were not a 
major influence.
Conflict and epidemics would most likely have had 
substantial effects on either or both of the new
time-varying variables included.

@@

Although we cannot entirely rule out that other, unobserved
factors play a role, our main conclusion is that households appear
to consciously delay fertility through contraceptive use to cope 
with income shocks. 
This is especially of interest since this is accomplished 
predominately through the use of traditional contraceptives.
If traditional contraceptives have low efficacy, a natural question is 
how the substantial reductions in likelihood of births and pregnancy 
come about.
We present suggestive evidence that shows that only those households
with a strong incentive, i.e. those exposed to crop loss, are 
effective users of traditional contraceptives.%
\footnote{
This is in line with the results in \cite{Rosenzweig1989},
where women with the highest opportunity cost of children
are also the most effective user of traditional contraceptives,
although the setting is very different.
}

There are three potential caveats to our results.
First, the sample size is small and the survey rounds
could have been further apart, but the longitudinal nature
of the data means that we can control for unobservable 
time-invariant individual and household characteristics.
Second, not surprisingly there appears to be measurement 
errors.
% Reported age and education levels are not always consistent
% across rounds, which is an indication that we probably also
% have measurement errors in the other variables, although it is
% harder to verify.
To the extent that these measurement errors are classical 
measurement errors, this biases our results towards zero,
making it harder to show an effect of crop loss.%
\footnote{
The problems with measurement error, combined with the small
sample size, may also explain why we do not find clear
differences in the response to crop loss across age groups,
education, and wealth levels.
}
Third, although we cover a number of possible explanations
for what determines crop loss, it is possible that there may 
be other unobserved, time-varying household characteristics 
that influence whether a household experience a crop loss.
If that is the case the causality from crop loss to fertility 
decisions may not hold.

@@

What implications do our results have for our understanding 
of long-term fertility?
The economic literature on fertility transition in Europe and the US 
show a substantial decline in fertility and control of the timing of 
fertility before the introduction of modern contraceptives, although 
the methods used are generally not well documented.%
\footnote{
See, for example, \citet{Guinnane2011} on fertility reduction and 
\citet{Cinnirella2017} on spacing of birth.
On contraceptive methods used see, for example, \citet{Michael1976},
\citet{David1986}, and \citet{Santow1995}.
}
Furthermore, studies of the introduction of modern family planning, 
using individual level data, rarely find reductions in fertility larger 
than one child.%
\footnote{
See, for example, \citet{Portner2014a} 
and \citet{Portner2018} and the literature surveyed therein.
}
Both set of results indicate that households are able to 
limit long-term fertility using traditional contraceptives.
Our individual level data allow us to directly link the use of
traditional contraceptives and fertility outcomes.
We argue that, contrary to common perception, traditional
contraceptive methods have a high degree of efficacy when the
incentives to prevent pregnancy or birth are sufficiently high.
Because of the small data set we cannot, however, differentiate between
the different traditional methods, although---contrary to Europe 
and the US---withdrawal plays little or no role in this area.

@@

Including only the additional lag of crop loss it appears that 
there is a weak positive association between crop loss in the prior
period and the likelihood of pregnancy and birth, although the 
coefficients are small and statistically insignificant.
There is still a negative effect of the immediate crop loss
variable, although these are also not statistically significant.
These results are broadly consistent with a compensatory response
where crop loss longer ago leads to higher fertility now.
However, if we include the interaction between crop losses we 
instead find negative association for both immediate and lagged 
crop loss with a positive interaction coefficient.
Even with the positive interaction coefficient the estimated 
impact of crop losses in the two prior periods is negative.
For contraceptive use we see similar patterns (with the signs 
reversed).
